---
title: "Preprocessing of mouse T-cell lymphoma"
author: "Katharina Schmid"
date: "2025-01-21"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning=FALSE)
knitr::opts_knit$set(root.dir = "..")

library(Seurat)
library(data.table)
library(Matrix)
library(ggplot2)
```

Data from UMCG (Floris and Andréa). Sample from the Thyme, should contain mostly T cells and a small fraction from B cells that
should be filtered out with the respective markers, also suggested by Andréa.

Then an euploid cancer sample is added additionally as a negative reference for the CNV callers. 

## Tumor data from aneuploid mouse

### Create Seurat object, quality control & filtering

```{r}

counts<-readRDS("data/mouse_UMCG/RNA_T989_SeekGene/T989_count_matrix.rds")

#Create a Seurat object
seurat_obj <- CreateSeuratObject(counts=counts,min.cells=3,min.features=200)

seurat_obj

```

```{r}

#Calculate the percentage of mitochondrial counts
seurat_obj[["percent.mt"]] <- PercentageFeatureSet(seurat_obj, pattern = "^mt-")
VlnPlot(seurat_obj, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3,
        layer="count")

```

```{r}
plot1 <- FeatureScatter(seurat_obj, feature1 = "nCount_RNA", feature2 = "percent.mt")
plot2 <- FeatureScatter(seurat_obj, feature1 = "nCount_RNA", feature2 = "nFeature_RNA")
plot1 + plot2
```

```{r}
ggplot(seurat_obj@meta.data,aes(x=nCount_RNA))+
  geom_histogram(bins=50)

ggplot(seurat_obj@meta.data,aes(x=nFeature_RNA))+
  #geom_vline(xintercept=500)+
  geom_histogram(bins=50)

```

```{r}

seurat_obj <- subset(seurat_obj, subset = percent.mt < 20 & nCount_RNA > 1000)
seurat_obj

```


```{r}
ggplot(seurat_obj@meta.data,aes(x=nCount_RNA))+
  geom_histogram()+
  geom_vline(xintercept=1000)

# ggplot(seurat_obj@meta.data,aes(x=nFeature_RNA))+
#   geom_histogram(bins=50)

```

### Normalize data

```{r}

seurat_obj<-NormalizeData(seurat_obj,normalization.method = "LogNormalize",
                          scale.factor=10000)


```

### Identify highly variable features

```{r,fig.height=4,fig.width=10}

seurat_obj <- FindVariableFeatures(seurat_obj, selection.method = "vst", nfeatures = 2000)

# Identify the 10 most highly variable genes
top10 <- head(VariableFeatures(seurat_obj), 10)

# plot variable features with and without labels
plot1 <- VariableFeaturePlot(seurat_obj)
plot2 <- LabelPoints(plot = plot1, points = top10, repel = TRUE)
plot1 + plot2

```

## PCA

```{r}

#Scale the data
all.genes <- rownames(seurat_obj)
seurat_obj <- ScaleData(seurat_obj, features = all.genes)

seurat_obj <- RunPCA(seurat_obj, features = VariableFeatures(object = seurat_obj),
                     verbose=FALSE)
DimPlot(seurat_obj, reduction = "pca")

```
```{r}
ElbowPlot(seurat_obj)
```

## Find clusters

```{r}
seurat_obj <- FindNeighbors(seurat_obj, dims = 1:15)
seurat_obj <- FindClusters(seurat_obj, resolution = 1)
```
## Plot UMAP

```{r}

seurat_obj <- RunUMAP(seurat_obj, dims = 1:10)
DimPlot(seurat_obj, reduction = "umap")

```
Show coverage:

```{r}
FeaturePlot(seurat_obj, features = c("nCount_RNA"))
```

## Annotate clusters based on marker genes

Cd3e for T cells

Cd74 & Itgax for B cells

```{r}
#GATA2 and FCER1A not expressed high enough in our dataset
used_marker_genes <- c("Cd3e", "Cd74","Itgax")
VlnPlot(seurat_obj, features = used_marker_genes)

FeaturePlot(seurat_obj, features = used_marker_genes)

```

```{r}

new.cluster.ids <- rep("T_cells", 11)
new.cluster.ids[12]<-"B_cells"
names(new.cluster.ids) <- levels(seurat_obj)
seurat_obj <- RenameIdents(seurat_obj, new.cluster.ids)
DimPlot(seurat_obj, reduction = "umap", label = TRUE, pt.size = 0.5) + NoLegend()

```

```{r}

used_marker_genes <- c("Cd3e", "Cd74","Itgax")

VlnPlot(seurat_obj, features = used_marker_genes)

```

## Tumor data from euploid mouse

Also filter and 

```{r}

counts_ep<-readRDS("data/mouse_UMCG/euploid_cancer_reference/eT_count_matrix.rds")

#Create a Seurat object
seurat_obj_ref <- CreateSeuratObject(counts=counts_ep,min.cells=3,min.features=200)

seurat_obj_ref

```
```{r}

#Calculate the percentage of mitochondrial counts
seurat_obj_ref[["percent.mt"]] <- PercentageFeatureSet(seurat_obj_ref, pattern = "^mt-")
VlnPlot(seurat_obj_ref, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3,
        layer="count")

```

```{r}
plot1 <- FeatureScatter(seurat_obj_ref, feature1 = "nCount_RNA", feature2 = "percent.mt")
plot2 <- FeatureScatter(seurat_obj_ref, feature1 = "nCount_RNA", feature2 = "nFeature_RNA")
plot1 + plot2
```

```{r}
ggplot(seurat_obj_ref@meta.data,aes(x=nCount_RNA))+
  geom_vline(xintercept=1000)+
  geom_histogram(bins=50)

ggplot(seurat_obj_ref@meta.data,aes(x=nFeature_RNA))+
  #geom_vline(xintercept=500)+
  geom_histogram(bins=50)

```


```{r}
seurat_obj_ref <- subset(seurat_obj_ref, subset = percent.mt < 20 & nCount_RNA > 1000)
seurat_obj_ref

```

## Combine both to one large count matrix for the CNV RNA callers


```{r}

#Extract the Seurat count matrices
counts_tumor<-as.matrix(seurat_obj[["RNA"]]$counts)
counts_ref<-as.matrix(seurat_obj_ref[["RNA"]]$counts)

#Find common gene set
common_genes<-intersect(rownames(counts_ref),rownames(counts_tumor))
print(paste("Identified",length(common_genes),"common genes!"))

#Filter and combine both
counts_tumor<-counts_tumor[common_genes,]
counts_ref<-counts_ref[common_genes,]
counts_combined<-cbind(counts_tumor,counts_ref)

#Save sample annotation
sample_annot<-data.frame(barcode=colnames(counts_combined),
                         sample=c(rep("mouse",ncol(counts_tumor)),
                                  rep("ref",ncol(counts_ref))))


write.table(sample_annot, file="data/mouse_UMCG/input_mouse/sample_annotation.txt",
            sep="\t",quote=FALSE,row.names=FALSE,col.names=FALSE)


df_refs<-data.frame(ref_groups=setdiff(unique(sample_annot$sample),"mouse"))
write.table(df_refs, file="data/mouse_UMCG/input_mouse/ref_groups.txt",
            sep="\t",quote=FALSE,row.names=FALSE,col.names=TRUE)

#Save the reduced count matrix
write.table(counts_combined,
            file="data/mouse_UMCG/input_mouse/count_matrix.txt",
            sep="\t",quote=FALSE,row.names=TRUE)

```

