---
title: "Prepare MM scRNAseq data"
output: html_document
date: "2023-03-06"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning=FALSE)
knitr::opts_knit$set(root.dir = "~/Documents/CNV_RNAseq_benchmark")

library(Seurat)
library(data.table)
library(Matrix)
library(ggplot2)
library(scran)

```

## Load dataset (skipped for runtime reasons)

Two different versions of the count matrix exist: the one from GEO and the one
created by us personally from cell ranger

```{r,eval=FALSE}
#Load GEO count matrix (actually splitted into two for some weird reason)
counts<-fread("data/MM_sciCNV/GSE141299_RAW/GSM4200471_MM199_All_Celltypes_RawData.txt.gz")
mito<-fread("data/MM_sciCNV/GSE141299_RAW/GSM4200471_MM199_Mitochondrial_Expression.txt.gz")

#Combine both into one count matrix
colnames(mito)[1]<-"gene.names"
counts<-rbind(counts,mito)
rm(mito)
gc()

#Convert it into a matrix
gene_names<-counts$gene.names
counts$gene.names<-NULL
counts<-as.matrix(counts)
rownames(counts)<-gene_names

#Load cellranger count matrix
cr_counts<-readMM("data/MM_sciCNV/cellranger_output/filtered_feature_bc_matrix/matrix.mtx.gz")
genenames<-read.table("data/MM_sciCNV/cellranger_output/filtered_feature_bc_matrix/features.tsv.gz")
cellnames<-read.table("data/MM_sciCNV/cellranger_output/filtered_feature_bc_matrix/barcodes.tsv.gz")

colnames(cr_counts)<-cellnames$V1
rownames(cr_counts)<-genenames$V1


# #Check how well the overall UMI counts match
# ov_counts_geo<-data.frame(barcodes=colnames(counts),
#                           counts_geo=colSums(counts))
# 
# ov_counts_cr<-data.frame(barcodes=colnames(cr_counts),
#                           counts_cr=colSums(cr_counts))
# ov_counts_cr$barcodes<-gsub("-1","",ov_counts_cr$barcodes)
# 
# ov_counts<-merge(ov_counts_geo,ov_counts_cr,by="barcodes")
# 
# ggplot(ov_counts,aes(x=counts_geo,y=counts_cr))+
#   geom_point()+
#   xlab("Counts per cell - GEO")+ylab("Counts per cell - CellRanger")
# 
# #Merge the counts per gene if multiple genes point to the same gene
# cr_counts_merged<-apply(cr_counts,2,tapply,as.factor(genenames$V2),sum,na.rm=T)
# 
# #Check how well the counts per gene match
# ov_counts_geo<-data.frame(gene=rownames(counts),
#                           counts_geo=rowSums(counts))
# 
# ov_counts_cr<-data.frame(gene=rownames(cr_counts_merged),
#                           counts_cr=rowSums(cr_counts_merged))
# 
# ov_counts<-merge(ov_counts_geo,ov_counts_cr,by="gene")
# 
# ggplot(ov_counts,aes(x=counts_geo,y=counts_cr))+
#   geom_point()+scale_y_log10()+scale_x_log10()+
#   xlab("Counts per gene - GEO")+ylab("Counts per gene - CellRanger")
# 
# rm(cr_counts_merged,cr_counts)
# gc()

```

Decided to go with cellranger matrix for analysis (probably no big difference in the end)

```{r}

#Load cellranger count matrix
counts<-readMM("data/MM_sciCNV/cellranger_output/filtered_feature_bc_matrix/matrix.mtx.gz")
genenames<-read.table("data/MM_sciCNV/cellranger_output/filtered_feature_bc_matrix/features.tsv.gz")
cellnames<-read.table("data/MM_sciCNV/cellranger_output/filtered_feature_bc_matrix/barcodes.tsv.gz")

colnames(counts)<-cellnames$V1
rownames(counts)<-genenames$V2

#Combine duplicate gene names to one row
counts<-apply(counts, 2, tapply, as.factor(genenames$V2),sum, na.rm=T)
  
#Create a Seurat object
seurat_obj <- CreateSeuratObject(counts=counts,min.cells=3,min.features=200)

seurat_obj

```

## Quality control

```{r}

#Calculate the percentage of mitochondrial counts
seurat_obj[["percent.mt"]] <- PercentageFeatureSet(seurat_obj, pattern = "^MT-")
VlnPlot(seurat_obj, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3,
        layer="count")

```

```{r}
plot1 <- FeatureScatter(seurat_obj, feature1 = "nCount_RNA", feature2 = "percent.mt")
plot2 <- FeatureScatter(seurat_obj, feature1 = "nCount_RNA", feature2 = "nFeature_RNA")
plot1 + plot2
```

```{r}
seurat_obj <- subset(seurat_obj, subset = nCount_RNA > 3000 & percent.mt < 5)

seurat_obj
```
## Normalize data

Now testing scran
Other tested options: logNormalize, scTransform

```{r}

sce <- SingleCellExperiment(list(counts=as.matrix(seurat_obj@assays$RNA@layers$counts)))
clusters <- quickCluster(sce)
sce <- computeSumFactors(sce, clusters=clusters)
sce <- logNormCounts(sce,log=FALSE)
log_counts<-log(normcounts(sce)+1)
seurat_obj[["RNA"]]$data<-as(log_counts,"dgCMatrix")

#seurat_obj<-SCTransform(seurat_obj,vars.to.regress = "percent.mt", verbose=TRUE)

# seurat_obj<-NormalizeData(seurat_obj,normalization.method = "LogNormalize",
#                           scale.factor=10000)

```


## Identify highly variable features

```{r,fig.height=4,fig.width=10}

seurat_obj <- FindVariableFeatures(seurat_obj, selection.method = "vst", nfeatures = 2000)

# Identify the 10 most highly variable genes
top10 <- head(VariableFeatures(seurat_obj), 10)

# plot variable features with and without labels
plot1 <- VariableFeaturePlot(seurat_obj)
plot2 <- LabelPoints(plot = plot1, points = top10, repel = TRUE)
plot1 + plot2

```

## Scaling the data

```{r}

all.genes <- rownames(seurat_obj)
seurat_obj <- ScaleData(seurat_obj, features = all.genes)

```

## PCA

```{r}
seurat_obj <- RunPCA(seurat_obj, features = VariableFeatures(object = seurat_obj),
                     verbose=FALSE)
DimPlot(seurat_obj, reduction = "pca")
```
```{r}
ElbowPlot(seurat_obj)
```
## Find clusters

```{r}
seurat_obj <- FindNeighbors(seurat_obj, dims = 1:15)
seurat_obj <- FindClusters(seurat_obj, resolution = 1)
```
## Plot UMAP

```{r}
seurat_obj <- RunUMAP(seurat_obj, dims = 1:10)
DimPlot(seurat_obj, reduction = "umap")
```


## Check known cancer marker genes

Specific MM marker genes: B2M, CCND1, Immunoglobulins (IGHM, IGKC)

Source: https://www.cancer.gov/about-cancer/diagnosis-staging/diagnosis/tumor-markers-list

Additional marker used in the Numbat paper for MM: MZB1

```{r}

VlnPlot(seurat_obj, features = c("B2M","CCND1","IGHM","IGKC","MZB1"))
FeaturePlot(seurat_obj, features = c("B2M","CCND1","IGHM","IGKC","MZB1"))

```

## Annotate based on the cancer marker genes

```{r}

#Define the two large clusters
seurat_obj$large_clusters<-ifelse(seurat_obj$RNA_snn_res.0.5 %in% c(4,5,7),
                                  "control","MM")

DimPlot(seurat_obj, reduction = "umap",group.by="large_clusters")

```

```{r,eval=FALSE}
cluster_markers <- FindMarkers(seurat_obj, ident.1 = seq(0,4), 
                                 min.pct = 0.25,
                               test.use="wilcox")

print(head(cluster_markers))

#Save all DE genes with p-value < 0.05
de_genes<-cluster_markers[cluster_markers$p_val_adj<0.05,]
write.table(de_genes,file="data/MM_sciCNV/de_genes_cancer.tsv",sep="\t",quote=FALSE)
print(nrow(de_genes))

# Top positive markers
top_pos_markers<-rownames(cluster_markers)[cluster_markers$avg_log2FC>0]
VlnPlot(seurat_obj, features = top_pos_markers[1:5])

FeaturePlot(seurat_obj, features=top_pos_markers[1:5])

#Top negative markers
top_neg_markers<-rownames(cluster_markers)[cluster_markers$avg_log2FC<0]
VlnPlot(seurat_obj, features = top_neg_markers[1:5])

FeaturePlot(seurat_obj, features=top_neg_markers[1:5])

```

```{r,eval=FALSE}

#Save sample annotation
meta_data<-seurat_obj@meta.data
sample_annot<-data.frame(barcode=rownames(meta_data),
                               sample=meta_data$large_clusters)

write.table(sample_annot, file="data/MM_sciCNV/input_MM/sample_annotation.txt",
            sep="\t",quote=FALSE,row.names=FALSE,col.names=FALSE)


df_refs<-data.frame(ref_groups=setdiff(unique(sample_annot$sample),"MM"))
write.table(df_refs, file="data/MM_sciCNV/input_MM/ref_groups.txt",
            sep="\t",quote=FALSE,row.names=FALSE,col.names=TRUE)

#Save the reduced count matrix
counts_reduced<-seurat_obj@assays$RNA@counts
write.table(as.matrix(counts_reduced),
            file="data/MM_sciCNV/input_MM/count_matrix.txt",
            sep="\t",quote=FALSE,row.names=TRUE)


```

## Compare annotation with results for SCEVAN and copyKat

```{r, fig.width=15}

meta_data<-seurat_obj@meta.data

#Read prediction files
input_copykat<-"code/benchmark_scrna_cnv_caller/snakemake_pipeline/results/output_MM/copykat_predict_cancer_cells/MM_copykat_prediction.txt"
pred_copykat <-fread(input_copykat)
#pred_copykat$cell.names<-paste0(pred_copykat$cell.names,"-1")

input_scevan<-"code/benchmark_scrna_cnv_caller/snakemake_pipeline/results/output_MM/scevan_predict_cancer_cells/output/MM_scevan_prediction.txt"
suppressWarnings(pred_scevan <-fread(input_scevan))
pred_scevan$V1<-paste0(pred_scevan$V1,"-1")
colnames(pred_scevan)[2]<-"scevan_pred"

#Add the information to the Seurat meta information
meta_data$cell_name<-rownames(meta_data)

meta_data<-merge(meta_data,pred_copykat,by.x="cell_name",by.y="cell.names",
                 all.x=TRUE,sort=FALSE)

meta_data<-merge(meta_data,pred_scevan,by.x="cell_name",by.y="V1",
                 all.x=TRUE,sort=FALSE)

seurat_obj$copykat_pred<-meta_data$copykat.pred
seurat_obj$scevan_pred<-meta_data$scevan_pred

print(table(seurat_obj$large_clusters,seurat_obj$copykat_pred))
print(table(seurat_obj$large_clusters,seurat_obj$scevan_pred))
print(table(seurat_obj$copykat_pred,seurat_obj$scevan_pred))

DimPlot(seurat_obj, reduction = "umap",group.by=c("copykat_pred","scevan_pred"))

```

