---
title: "Check annotation problems iAMP21"
author: "Katharina Schmid"
date: "2025-01-29"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "..")

library(data.table)
library(Seurat)
```



## Check the annotations from CopyKat and SCEVAN

Problem: both CopyKat and SCEVAN fail to annotate the cells properly

```{r}

seurat_obj<-readRDS("data/iAMP21/seurat_obj.RDS")

#Percentage of cancer cells
print(round(mean(Idents(seurat_obj)=="Blast"),3))

#Percentage of non-cancer cells
print(round(mean(Idents(seurat_obj)!="Blast"),3))

```

### Check CopyKat

```{r}

copykat_pred<-fread("snakemake_pipeline/results/output_iAMP21/copykat_predict_cancer_cells/iAMP21_copykat_prediction.txt")
table(copykat_pred$copykat.pred)

seurat_obj$copykat_pred<-"filtered"
seurat_obj$copykat_pred[copykat_pred$cell.names]<-copykat_pred$copykat.pred

table(Idents(seurat_obj),seurat_obj$copykat_pred)

```

```{r}

DimPlot(seurat_obj, reduction = "umap", group.by="copykat_pred")

```

### Check SCEVAN

```{r}

scevan_pred<-fread("snakemake_pipeline/results/output_iAMP21/scevan_predict_cancer_cells/output/iAMP21_scevan_prediction.txt")
table(copykat_pred$class)

seurat_obj$scevan_pred<-"filtered"
seurat_obj$scevan_pred[scevan_pred$V1]<-scevan_pred$class

table(Idents(seurat_obj),seurat_obj$scevan_pred)

```


```{r}

DimPlot(seurat_obj, reduction = "umap", group.by="scevan_pred")

```


Extract the confidental normal cells from SCEVAN defined as seed for the analysis:

```{r}
res_proc<-readRDS("snakemake_pipeline/results/output_iAMP21/scevan_predict_cancer_cells/output/first_healthy_cell_annotation.RDS")

table(res_proc$norm_cell)

seurat_obj$scevan_normal<-"no"
seurat_obj$scevan_normal[names(res_proc$norm_cell)]<-res_proc$norm_cell

table(Idents(seurat_obj),seurat_obj$scevan_normal)  

DimPlot(seurat_obj, reduction = "umap")

DimPlot(seurat_obj, reduction = "umap", 
        cells.highlight	=list("Bcell"=names(res_proc$norm_cell)[res_proc$norm_cell=="Bcells"],
                              "macro"=names(res_proc$norm_cell)[res_proc$norm_cell=="macro"],
                              "Tcell"=names(res_proc$norm_cell)[res_proc$norm_cell=="Tcell"]),
        cols.highlight =c("darkred","darkgreen","darkblue") )

```

## Check the other three datasets - how did the initial annotation work for them

### MM dataset

```{r}

dataset<-"MM"
res_proc<-readRDS(paste0("snakemake_pipeline/results/output_",dataset,
                         "/scevan_predict_cancer_cells/output/first_healthy_cell_annotation.RDS"))

table(res_proc$norm_cell)

annot<-fread(paste0("snakemake_pipeline/data/input_",dataset,"/sample_annotation.txt"),header=FALSE)
annot<-as.data.frame(annot)
rownames(annot)<-annot$V1

annot$scevan_normal<-"no"
annot[names(res_proc$norm_cell),]$scevan_normal<-res_proc$norm_cell

annot[annot$scevan_normal != "no",]

```
### BCC06 dataset

BCC06: found no confident normal cells, used a synthetic baseline

```{r}

dataset<-"BCC06"
res_proc<-readRDS(paste0("snakemake_pipeline/results/output_",dataset,
                         "/scevan_predict_cancer_cells/output/first_healthy_cell_annotation.RDS"))

table(res_proc$norm_cell)

# annot<-fread(paste0("snakemake_pipeline/data/input_",dataset,"/sample_annotation.txt"),header=FALSE)
# annot<-as.data.frame(annot)
# rownames(annot)<-annot$V1
# 
# annot$scevan_normal<-"no"
# annot[names(res_proc$norm_cell),]$scevan_normal<-res_proc$norm_cell
# 
# annot[annot$scevan_normal != "no",]

```

### BCC06post dataset

```{r}

dataset<-"BCC06post"
res_proc<-readRDS(paste0("snakemake_pipeline/results/output_",dataset,
                         "/scevan_predict_cancer_cells/output/first_healthy_cell_annotation.RDS"))

table(res_proc$norm_cell)

annot<-fread(paste0("snakemake_pipeline/data/input_",dataset,"/sample_annotation.txt"),header=FALSE)
annot<-as.data.frame(annot)
rownames(annot)<-annot$V1

annot$scevan_normal<-"no"
annot[names(res_proc$norm_cell),]$scevan_normal<-res_proc$norm_cell

table(annot$V2,annot$scevan_normal)

```